### 2016-09-13 paper/algorithms.tex
stashing some paper changes

\section{Inference Algorithm and Complexity Analysis}\label{sect:alg}
%
Based on the theoretical developments in the last section, we present below a goal-oriented inference algorithm that allows one to combine models of varying fidelity, while maintaining rigorous control of QoI error.
%
\subsection{Goal Oriented Inference Algorithm}
%
Just as error estimates can be used to guide mesh-refinement~\cite{BecRann01}, the error estimate \cref{eq:finErrExp} can be localized to give elemental contributions and used to guide the division of the domain for a mixed-fidelity model. The error estimate can be calculated again, using the mixed-fidelity model as the lower-fidelity model. This process can be repeated, successively increasing the proportion of the domain in which the high-fidelity model is used, until some threshold is reached. Algorithm~\cref{alg:refSeries} describes this approach. Note that the QoI error adjoint problem \cref{eq:superAdjEq} involves linearization about $\Psi_{HF}$, which is not available, so in the case of a nonlinear goal-oriented inverse problem, the QoI error adjoint problem is approximated by linearizing about $\Psi_{LF}$ instead.
%
\alglanguage{pseudocode}
\begin{algorithm}[h!]
\small
\caption{An algorithm to adaptively build a mixed-fidelity model for low error in the QoI.}
\label{alg:refSeries}
\begin{algorithmic}[1]
\State{Define maximum acceptable absolute relative QoI error \texttt{errTol}}
\State{Define maximum number of adaptive iterations \texttt{maxIter}}
\Procedure{$\texttt{BuildMF}$}{HF model, LF model, \texttt{errTol}, \texttt{maxIter}}
	\State{Let the model MF$_0$ be the LF model applied everywhere in the domain.}
	\State{$i\gets0$}
	\State{Solve for stationary point $\Psi_{MF_0}$ of augmented Lagrangian $\mathcal{M}_{MF_0}$}
	\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_0}$, for \par\hskip\algorithmicindent supplementary adjoint $\Lambda_0$ (see \cref{eq:superAdjEq})}
	\State{Compute QoI error estimate

	$\epsilon_0=-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{MF_0})(\Lambda_0)+\mathcal M_{HF}(\Psi_{MF_0})-\mathcal M_{MF_0}(\Psi_{MF_0})$}
	\State{Calculate QoI $I(q_{MF_0},u_{MF_0})$}
	\While{$i<$ \texttt{maxIter} and $|\epsilon_i/I(q_{MF_i},u_{MF_i})|>$ \texttt{errTol}}
		\State{\begin{varwidth}[t]{\linewidth}Localize $\epsilon_i$ (see \cref{sec:errLocal}) and use this decomposition to guide \par\hskip\algorithmicindent formation of new mixed-fidelity model MF$_{i+1}$\end{varwidth}}
		\State{$i\gets i+1$}
		\State{Solve for stationary point $\Psi_{MF_i}$ of augmented Lagrangian $\mathcal{M}_{MF_i}$}
		\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_i}$, for

@@

\Cref{alg:refSeries} can be naturally extended to an online-offline setting, where a multi-fidelity model adaptive constraint can be generated for a given data set in the offline phase. In the online phase, additional data points are added to the data set, and instead of solving the complete inverse problem with all the data points, and the high-fidelity constraints, an iterative algorithm can be initiated using the multi-fidelity representation developed in the offline phase. The adaptive procedure, with the supplementary adjoint can be used to adapt the preexisting multi-fidelity representation to the new data points.  
%
\alglanguage{pseudocode}
\begin{algorithm}[h!]
\small
\caption{An algorithm to adapt a pre-computed mixed-fidelity model to new data.}
\label{alg:refonoff}
\begin{algorithmic}[1]
\State{Define maximum acceptable absolute relative QoI error \texttt{errTol}}
\State{Given a data set $d_1 \in \R^{n_{d_1}}$, and an adaptively build a mixed-fidelity model for low error in the QoI, use \cref{alg:refSeries} to build a MF model MF$_1$} 
\Procedure{$\texttt{Online}$}{HF model, MF model MF$_1$, \texttt{errTol}, $d_2 \in \R^{n_{d_2}}$}
	\State{$i\gets1$}
	\State{Solve for stationary point $\Psi_{MF_1}$ of augmented Lagrangian $\mathcal{M}_{MF_1}$ formed using new data set $d_2$}
	\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_1}$, for \par\hskip\algorithmicindent supplementary adjoint $\Lambda_1$ (see \cref{eq:superAdjEq})}
	\State{Compute QoI error estimate

	$\epsilon_0=-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{MF_1})(\Lambda_1)+\mathcal M_{HF}(\Psi_{MF_1})-\mathcal M_{MF_1}(\Psi_{MF_1})$}
	\State{Calculate QoI $I(q_{MF_1},u_{MF_1})$}
	\While{$i<$ \texttt{maxIter} and $|\epsilon_i/I(q_{MF_i},u_{MF_i})|>$ \texttt{errTol}}
		\State{\begin{varwidth}[t]{\linewidth}Localize $\epsilon_i$ (see \cref{sec:errLocal}) and use this decomposition to guide \par\hskip\algorithmicindent formation of new mixed-fidelity model MF$_{i+1}$\end{varwidth}}
		\State{$i\gets i+1$}
		\State{Solve for stationary point $\Psi_{MF_i}$ of augmented Lagrangian $\mathcal{M}_{MF_i}$}
		\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_i}$, for

		$\quad\quad \epsilon_i=-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{MF_i})(\Lambda_i)+\mathcal M_{HF}(\Psi_{MF_i})-\mathcal M_{MF_i}(\Psi_{MF_i})$}
		\State{Calculate QoI $I(q_{MF_i},u_{MF_i})$}
	\EndWhile \\
\Return{QoI estimate $I(q_{MF_i},u_{MF_i})$ for the new data set $d_2$}
\EndProcedure
\Statex
\end{algorithmic}
\end{algorithm}
%
\Cref{alg:refonoff} lends itself to a quasi-incremental data assimilation approach, where the data set $d_2$ is an augmentation of the original set $d_1$. In such a situation, we expect that the model refinement will be concentrated around the new data points, that also inform the QoI. Thus, in the online phase, even though the inverse problem will be solved with all data points, the use of high-fidelity models will be limited, and the bulk of the computation (computing the auxilary variables, and supplementary adjoint) will be focused on identifying those new data points, which have the maximum impact on the QoI.

\Cref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space. The two models could also correspond to two levels of mesh-refinement, though in this case the method described in~\cite{BecVex05} could be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF,\Psi}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.
%------------------------------------------------------------------------------------------------------------------------%
\subsubsection{Error Estimate Localization}\label{sec:errLocal}
%------------------------------------------------------------------------------------------------------------------------%
Algorithm~\cref{alg:refSeries} does not require a specific method for localizing the error estimate. A na\"{i}ve approach would be to write the error estimate as a sum of integrals over elements and their boundaries, and calculate the error contribution by each element as the integral over that element. While simple, this method can lead to non-zero error contributions from elements in which the high-fidelity model is already being used, making the error decomposition more difficult to interpret and use for refinement.

We instead use the alternative method described in \cite{vanOpstaletal15}, decomposing the error estimate into contributions from locally supported basis functions rather than elements. For convenience of notation in this section, we drop subscripts so that $\Psi=\Psi_{LF}$, $Q=Q_{HF}$, and $U=U_{HF}$; for our numerical experiments, we have $Q_{LF}\subseteq Q$ and $U_{LF}=U$. Note that the error estimate in \cref{eq:finErrExp} can be equivalently written as,
%
\begin{equation}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi)(\Lambda)+\mathcal{L}'_{HF,\xi}(\xi_\Psi)(\chi_\Psi)+\mathcal{R}(e^3). \nonumber
\end{equation}
%
We consider a finite-dimensional \red{conforming} subspace $(Q^h\times U^h\times U^h)^3 \subset (Q\times U\times U)^3$ which contains the approximations $(\Lambda^h,\chi_\Psi^h)$. Define a basis $\Phi^h=\{(\varphi,\upvarphi)_i\}_{i\in I}$ consisting of locally supported functions such that span $\Phi^h=(Q^h\times U^h\times U^h)^3$; we can then write $(\Lambda^h,\chi_\Psi^h)=(\sum\limits_{i\in I}\varphi_i\lambda_i,\sum\limits_{i\in I}\upvarphi_i \chi_i)$. The error estimate,
%
\begin{equation}
\epsilon = -\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi)(\Lambda)+\mathcal{L}'_{HF,\xi}(\xi_\Psi)(\chi_\Psi) \leq \sum_{i\in I} \varepsilon_i,
\end{equation}
%
where,
%
\begin{equation}\label{eq:basisblame}
\varepsilon_i = \left| -\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi)(\lambda_i\varphi_i)+\mathcal{L}'_{HF,\xi}(\xi_\Psi)(x_i\upvarphi_i) \right|
\end{equation}
%
can be interpreted as the error contribution from the basis function $(\varphi,\upvarphi)_i$. Near the interfaces between the low-fidelity and high-fidelity regions, basis functions may have their support divided between the two regions and thus have a nonzero error contribution. The basis functions with the largest error contributions are flagged, and the elements in their support are refined.  
%------------------------------------------------------------------------------------------------------------------------%
\subsection{Complexity Analysis}\label{sect:alg_complexity}
%------------------------------------------------------------------------------------------------------------------------%
We now derive the computational complexity of using Algorithm~\cref{alg:refSeries} to solve an inference problem, and compare it to extant strategies. Let the state, adjoint, and parameter variables each have $N$ degrees of freedom, and assume that an `all at once' strategy is used to solve the KKT system. The cost of single linear solve with such a system is $\orderof{3N}^{\gamma}$, where $\gamma$ depends on the linear solver used~\cite{}. The cost of solving for the auxiliary variables, a linear system, is $\orderof{3N}^{\gamma}$. \red{The parts of the supplementary adjoint corresponding to the primary and auxiliary variables can be solved separately, and thus the cost of computing the supplementary adjoint is also $\orderof{3N}^{\gamma}$.}

If one utilizes Algorithm~\cref{alg:refSeries} to solve the inference problem, with $T$ adaptive iterations the total computational cost is,
%
\begin{align}
\label{eq:cost_adapt}
C_{MF} &= \sum_{i=1}^{T}L_{i}(3N)^{\gamma} + 1(3N)^{\gamma} + 2(3N)^{\gamma} \nonumber \\
&= \sum_{i=1}^{T} (L_i + 3) (3N)^{\gamma}
\end{align}
% 
where $L_i$ is the number of nonlinear solver steps needed to solve the KKT system for the low or mixed fidelity model at each adaptive step. The corresponding cost for solving the high fidelity inference problem, with a nonadaptive algorithm, based on a continuation scheme is,
%
\begin{equation}
\label{eq:cost_nadapt}
C_{HF} = \sum_{i=1}^{C}K_{i}(3N)^{\gamma}
\end{equation}
% 
where $K_i$ is the number of nonlinear solver steps needed to solve the KKT system for the high fidelity model at each continuation step, and $C$ is the number of continuation steps.

Comparing Eq.~\eqref{eq:cost_adapt} and~\eqref{eq:cost_nadapt}, we have,
%
\begin{equation}
\label{eq:cost_compare_0}
C_{MF} < C_{HF} \implies \sum_{i=1}^{T} (L_i + 3) < \sum\limits_{i=1}^{C}K_{i}
\end{equation}
%
If the low and multi-fidelity models are linear, or nonlinear in a very small region, then $L_i \approx 1$, and we have the following condition,
%
\begin{equation}
\label{eq:cost_compare_1}
T < \frac{\sum\limits_{i=1}^{C}K_{i}}{4},
\end{equation}
% 
for the adaptive algorithm to be less costly than solving the high-fidelity inverse problem. \red{I'm not sure this section makes sense, since this doesn't seem to hold for any of the actual runs we've done...our mixed models take fewer nonlinear steps than the HF model, but not close enough to one...}
%The bound on $T$ will be strictest if Gaussian elimination is used, with $\gamma = 3$, giving $T < \frac{\sum\limits_{i=1}^{C}K_{i}}{10}$.
%

>>>>

\section{Goal Oriented Inference Algorithm}\label{sect:alg}
%
Based on the theoretical developments in the last section, we now give a goal-oriented inference algorithm that allows one to combine models of varying fidelity, while maintaining rigorous control of QoI error.

Just as error estimates can be used to guide mesh-refinement~\cite{BecRann01}, the error estimate \cref{eq:finErrExp} can be localized to give elemental contributions and used to guide the division of the domain for a mixed-fidelity model. The error estimate can be calculated again, using the mixed-fidelity model as the lower-fidelity model. This process can be repeated, successively increasing the proportion of the domain in which the high-fidelity model is used, until some threshold is reached. Algorithm~\cref{alg:refSeries} describes this approach. Note that the QoI error adjoint problem \cref{eq:superAdjEq} involves linearization about $\Psi_{HF}$, which is not available, so in the case of a nonlinear goal-oriented inverse problem, the QoI error adjoint problem is approximated by linearizing about $\Psi_{LF}$ instead.
%
\alglanguage{pseudocode}
\begin{algorithm}[h!]
\small
\caption{An algorithm to adaptively build a mixed-fidelity model for low error in the QoI.}
\label{alg:refSeries}
\begin{algorithmic}[1]
\State{Define maximum acceptable absolute relative QoI error \texttt{errTol}}
\State{Define maximum number of adaptive iterations \texttt{maxIter}}
\Procedure{$\texttt{BuildMF}$}{HF model, LF model, \texttt{errTol}, \texttt{maxIter}}
	\State{Let the model MF$_0$ be the LF model applied everywhere in the domain.}
	\State{$i\gets0$}
	\State{Solve for stationary point $\Psi_{MF_0}$ of augmented Lagrangian $\mathcal{M}_{MF_0}$}
	\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_0}$, for \par\hskip\algorithmicindent supplementary adjoint $\Lambda_0$ (see \cref{eq:superAdjEq})}
	\State{Compute QoI error estimate

	$\epsilon_0=-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{MF_0})(\Lambda_0)+\mathcal M_{HF}(\Psi_{MF_0})-\mathcal M_{MF_0}(\Psi_{MF_0})$}
	\State{Calculate QoI $I(q_{MF_0},u_{MF_0})$}
	\While{$i<$ \texttt{maxIter} and $|\epsilon_i/(\epsilon_i+I(q_{MF_i},u_{MF_i}))|>$ \texttt{errTol}}
		\State{\begin{varwidth}[t]{\linewidth}Localize $\epsilon_i$ (see \cref{sec:errLocal}) and use this decomposition to guide \par\hskip\algorithmicindent formation of new mixed-fidelity model MF$_{i+1}$\end{varwidth}}
		\State{$i\gets i+1$}
		\State{Solve for stationary point $\Psi_{MF_i}$ of augmented Lagrangian $\mathcal{M}_{MF_i}$}
		\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_i}$, for

@@

\Cref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space~\cite{thesis/SIAM talk?}. The two models could also correspond to two levels of mesh-refinement, though in this case the method described in~\cite{BecVex05} could be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF,\Psi}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.

%------------------------------------------------------------------------------------------------------------------------%
\subsection{Extension to Offline-Online Setting}\label{sec:onoff}
%------------------------------------------------------------------------------------------------------------------------%

\Cref{alg:refSeries} can be naturally extended to an offline-online setting. One variation would be analagous to the offline-online decomposition proposed in \cite{LiebWill13}. The offline phase would consist of adaptively creating a mixed-fidelity model with an appropriate error tolerance given some expected observations $d^*$. When new data is received, one can then solve the inverse problem with the chosen mixed-fidelity model and the new data, and, if desired, compute an error estimate for the QoI. The mixed-fidelity inverse problems can be expected to require less time to solve than the high-fidelity inverse problems.

Another variation occurs in the context of a quasi-incremental data assimilation approach. A multi-fidelity model adaptive constraint can be generated for a given dataset $d_1$ in the offline phase. In the online phase, additional data points are included to form an augmented dataset $d_2$, and instead of solving the complete inverse problem with all the data points and the high-fidelity constraints, an iterative algorithm can be initiated using the multi-fidelity representation developed in the offline phase. The adaptive procedure, with the supplementary adjoint can be used to adapt the preexisting multi-fidelity representation to the new data points. In such a situation where $d_1\in d_2$, we expect that the model refinement will be concentrated around those of the new data points that also inform the QoI. Thus, in the online phase, even though the inverse problem will be solved with all data points $d_2$, the use of high-fidelity models will be limited, and the bulk of the computation (computing the auxilary variables, and supplementary adjoint) will be focused on identifying those new data points, which have the maximum impact on the QoI.

%------------------------------------------------------------------------------------------------------------------------%
\subsection{Error Estimate Localization}\label{sec:errLocal}
%------------------------------------------------------------------------------------------------------------------------%
Algorithm~\cref{alg:refSeries} does not require a specific method for localizing the error estimate. A na\"{i}ve approach would be to write the error estimate as a sum of integrals over elements and their boundaries, and calculate the error contribution by each element as the integral over that element. While simple, this method can lead to non-zero error contributions from elements in which the high-fidelity model is already being used, making the error decomposition more difficult to interpret and use for refinement.

We instead use the alternative method described in \cite{vanOpstaletal15}, decomposing the error estimate into contributions from locally supported basis functions rather than elements. For convenience of notation in this section, we drop subscripts so that $\Psi=\Psi_{LF}$, $Q=Q_{HF}$, and $U=U_{HF}$; for our numerical experiments, we have $Q_{LF}\subseteq Q$ and $U_{LF}=U$. Note that the error estimate in \cref{eq:finErrExp} can be equivalently written as,
%
\begin{equation}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi)(\Lambda)+\mathcal{L}'_{HF,\xi}(\xi_\Psi)(\chi_\Psi)+\mathcal{R}(e^3). \nonumber
\end{equation}
%
We consider a finite-dimensional \red{conforming} subspace $(Q^h\times U^h\times U^h)^3 \subset (Q\times U\times U)^3$ which contains the approximations $(\Lambda^h,\chi_\Psi^h)$. Define a basis $\Phi^h=\{(\varphi,\upvarphi)_i\}_{i\in I}$ consisting of locally supported functions such that span $\Phi^h=(Q^h\times U^h\times U^h)^3$; we can then write $(\Lambda^h,\chi_\Psi^h)=(\sum\limits_{i\in I}\varphi_i\lambda_i,\sum\limits_{i\in I}\upvarphi_i \chi_i)$. The error estimate,
%
\begin{equation}
\epsilon = -\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi)(\Lambda)+\mathcal{L}'_{HF,\xi}(\xi_\Psi)(\chi_\Psi) \leq \sum_{i\in I} \varepsilon_i,
\end{equation}
%
where,
%
\begin{equation}\label{eq:basisblame}
\varepsilon_i = \left| -\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi)(\lambda_i\varphi_i)+\mathcal{L}'_{HF,\xi}(\xi_\Psi)(x_i\upvarphi_i) \right|
\end{equation}
%
can be interpreted as the error contribution from the basis function $(\varphi,\upvarphi)_i$. Near the interfaces between the low-fidelity and high-fidelity regions, basis functions may have their support divided between the two regions and thus have a nonzero error contribution. The basis functions with the largest error contributions are flagged, and the elements in their support are refined.
